---
title: "Project 3"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  html_notebook: default
  word_document: default
---

#Introduction

This is Project 3 for STAT 557 2018 Spring by Meridith Bartley and Fei Jiang. The aim of this project is to implement a tree structured classifier using the splitting method in CART and a chosen split stopping criterion and then apply this classifier to a data set. 

#Description of Data

This car evaluation dataset is developed by Marko Bohanec and Blaz Zupan (1997). The response variable is the condition of a car which has four classes: unacceptable, acceptable, good, very good. There are six predictors to develop the model: buying price, price of the maintenance, number of doors, capacity in terms of persons to carry, the size of luggage boot, and estomated safety of the car. #https://rpubs.com/minma/cart_with_rpart# 



```{r packages to load, include=FALSE}
library(dplyr)
library(ggplot2)
library(flexclust)
library(grid)
library(gridExtra)
library(magrittr)
library(cluster)
library(fpc)
library(FNN)
```

```{r functions, include=FALSE}
p3data = read.csv("car.csv", header=F) 


```
## Exploritory Data Analysis

#Analysis


 





# Conclusions 



# Contributions 

The different tasks required to complete this project were equally divided between Meridith and Fei.Both members of this group contributed to this report. 
